Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv('pseudo_facebook.tsv', sep='\t')
head(pf, 10)
str(pf) # Get the structure of data
dim(pf) # Get the size
names(pf) #Get all columns
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
install.packages('ggplot2')
library(ggplot2)

qplot(x=dob_day, data=pf) +
    scale_x_continuous(breaks=1:31) + #was scale_x_discrete actually
    facet_wrap(~dob_month, ncol=4)

theme_set(theme_minimal(24)) 
ggplot(aes(x = dob_day), data = pf) + 
  geom_histogram(binwidth = 1) + 
  scale_x_continuous(breaks = 1:31)

ggplot(data = pf, aes(x = dob_day)) + 
  geom_histogram(binwidth = 1) + 
  scale_x_continuous(breaks = 1:31) + 
  facet_wrap(~dob_month)

```

***

#### What are some things that you notice about this histogram?
Response:

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}
ggplot(data = pf, aes(x = dob_day)) + 
  geom_histogram(binwidth = 1) + 
  scale_x_continuous(breaks = 1:31) + 
  facet_wrap(~dob_month)
qplot(x=dob_day, data=pf) +
    scale_x_continuous(breaks=1:31) + #was scale_x_discrete actually
    facet_wrap(~dob_month, ncol=4)
```

#### Let's take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x=friend_count, data=pf) #long tail 
qplot(x=friend_count, data=pf, xlim=c(0,1000)) 
qplot(x=friend_count, data=pf) +
    scale_x_continuous(limits=c(0,1000)) +
    scale_y_continuous(limits=c(0,15000)) #limit y axes if wanted

ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram()
```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

### Exploring with Bin Width
Notes:

```{r Limiting the Axes}
qplot(x=friend_count, data=pf, binwidth=25) +
    scale_x_continuous(limits=c(0,1000), breaks=seq(0,1000,50))
#http://docs.ggplot2.org/current/scale_continuous.html

ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))

```

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000), 
                     breaks = seq(0, 1000, 50)) +
    facet_wrap(~gender)

qplot(x = friend_count, data = pf) + 
  facet_grid(gender ~ .) 

ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
#missing values are NA by R
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000), 
                     breaks = seq(0, 1000, 50)) +
    facet_wrap(~gender)

```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x=tenure/365, data=pf, color=I('black'), fill = I('orange'), binwidth=0.25) +
    scale_x_continuous(breaks=seq(1,7,1), limits=c(0,7))

ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = .25, color = 'black', fill = '#F79420')

```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x=tenure/365, data=pf, 
      xlab='Number of years using Facebook',
      ylab='Number of users in sample',
      color=I('black'), fill = I('orange'), binwidth=0.25) +
    scale_x_continuous(breaks=seq(1,7,1), limits=c(0,7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}

```

***

### User Ages
Notes:

```{r User Ages}
qplot(x=age, data=pf, 
      xlab='Age',
      ylab='Number of users in sample',
      color=I('black'), fill = I('orange'), binwidth=1) +
    scale_x_continuous(breaks=seq(0,113,20), limits=c(0,150))


ggplot(aes(x = age), data = pf) + 
  geom_histogram(binwidth = 1, fill = '#5760AB') + 
  scale_x_continuous(breaks = seq(0, 113, 5))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:
```{r User Ages}
summary((pf$friend_count))
summary(log10(pf$friend_count+1)) #for 0 friend count, log of 0 is -ve infinity, hance add 1 as log 1 is 0
summary(sqrt(pf$friend_count))

qplot(x=friend_count, data=pf, 
      xlab='Friend Count',
      ylab='Number of users in sample',
      color=I('black'), fill = I('orange'), binwidth=100) +
    scale_x_continuous(breaks=seq(0,5000,500), limits=c(0,5000))

qplot(x=log10(friend_count+1), data=pf, 
      xlab='Friend Count',
      ylab='Number of users in sample',
      color=I('black'), fill = I('orange'), binwidth=100) +
    scale_x_log10(breaks=seq(0,5000,500), limits=c(0,5000))

library(gridExtra) 
p1 <- qplot(x=friend_count, data=pf)
p2 <- qplot(x=log10(friend_count+1), data=pf)
p3 <- qplot(x=sqrt(friend_count), data=pf)
grid.arrange(p1, p2, p3, ncol=1)

pgg1 <- ggplot(aes(x=friend_count), data=pf) + geom_histogram()
pgg2 <- pgg1 + scale_x_log10()
pgg3 <- pgg1 + scale_x_sqrt()
grid.arrange(pgg1, pgg2, pgg3, ncol=1)
```
***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
qplot(x=friend_count, data=pf) +
    scale_x_log10()
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, y= ..count../sum(..count..),
      xlab = 'Friend count',
      ylab = 'Proportion of Users with that friend count', 
      data = subset(pf, !is.na(gender)), binwidth = 10, geom='freqpoly', color=gender) +
  scale_x_continuous(limits = c(0, 1000), 
                     breaks = seq(0, 1000, 50))
#count vs density (plots percentages within each group)
ggplot(aes(x = friend_count, y = ..density../sum(..density..)), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot( x = www_likes, data = subset(pf, !is.na(gender)),
    geom='freqpoly', color=gender) +
    scale_x_log10()
summary(pf$www_likes)
table(pf$www_likes)
by(pf$www_likes, pf$gender, sum)

ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_log10()
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x=gender, y=friend_count, 
    data=subset(pf, !is.na(gender)), geom='boxplot'
    , ylim = c(0, 1000))

qplot(x=gender, y=friend_count, 
    data=subset(pf, !is.na(gender)), geom='boxplot') +
    scale_y_continuous(limits=c(0,1000)) #ignores outliers

qplot(x=gender, y=friend_count, 
    data=subset(pf, !is.na(gender)), geom='boxplot') +
    coord_cartesian(ylim=c(0,250)) #better than scale_y_continous as it doesn't ignore outliers

by (pf$friend_count, pf$gender, summary)
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x=gender, y=friend_count, 
    data=subset(pf, !is.na(gender)), geom='boxplot') +
    scale_y_continuous(limits=c(0,1000)) #ignores outliers
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
by (pf$friendships_initiated, pf$gender, summary)


qplot(x=gender, y=friendships_initiated, 
    data=subset(pf, !is.na(gender)), geom='boxplot') +
    coord_cartesian(ylim=c(0,150))

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
names(pf)
by(pf$friendships_initiated, pf$gender, summary)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary (pf$mobile_likes)
summary (pf$mobile_likes>0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

#The sum() function will not work since mobile_check_in is a factor variable. 
#You can use the length() function to determine the number of values in a vector.
#We could have also made mobile_check_in to hold boolean values. The sum()function does work on booleans (true is 1, false is 0).
sum(pf$mobile_check_in==1)/length(pf$mobile_check_in)

```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!