Exploring Prosper Loan Data by MNSHR
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
```

```{r echo=FALSE, Load_the_Data}
setwd("C:/Users/Manish/workspace/Udacity-DA/P4\ R")
getwd()
# Load the Data
pf <- read.csv('./prosperLoanData.csv', sep=',')
head(pf, 10)
str(pf) # Get the structure of data
dim(pf) # Get the size
names(pf) #Get all columns

ggplot(aes(x = LoanOriginalAmount), data = pf) + 
  geom_histogram(binwidth = 100, fill = '#5760AB') + 
  scale_x_continuous(breaks = seq(0, 113, 5))

```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}

```
####  Tip When plotting on a log scale, it is useful to note that 3 is about halfway between 1 and 10. As a side note, try not to plot counts on a log scale since counts of 0 are undefined and counts of 1 have a value of 0 (no height).
##### Tip: You can change the height and width of plots in code chunks with the fig.height and fig.width parameters in the chunk options.

# Univariate Analysis

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?


# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

#### Tip: Be mindful of the number of data points and variables that you put in a correlation matrix or plot matrix: you do not need to include all variables. In addition, you can use other packages not introduced in the associated course to conduct your exploration. Make sure you load them at the beginning of your document so that it is easiest to see which packages are necessary. (The above plot matrix comes from the psych package.)

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

#### Tip: Even when doing exploration, it can be good to select appropriate color palettes and set plot themes in order to make plots more readable. (The above plots use sequential color palettes from the RColorBrewer package; other variables might require qualitative or diverging palettes.)

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

#### Tip: Note that performing statistical tests and creating data models are not a required component of the project. Below, the mtable() call uses 'sdigits = 3' to make sure that three digits are printed after the decimal point in the summary statistics.

------

# Final Plots and Summary
#### Tip: Polish up the plots that you explored earlier in the report in the final plots section. Make sure you label and title the plots, and provide a description of what can be observed in each of the chosen plots.
### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
