Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:
Shape is 4 in stat summary, hence averages are marked by x. Male users are 
younger. 
In 2n plot, we can see that median friend count is larger for women. 
The estimates for old users > 70 is very noisy

```{r Third Qualitative Variable}
library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y=mean, geom='point', shape =5 )

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color=gender), stat='summary', fun.y=median)
```

***

### Plotting Conditional Summaries
Notes:
# Write code to create a new data frame,
# called 'pf.fc_by_age_gender', that contains
# information on each age AND gender group.

# The data frame should contain the following variables:

#    mean_friend_count,
#    median_friend_count,
#    n (the number of users in each age and gender grouping)

```{r Plotting Conditional Summaries}
library(dplyr)
age_gender_groups <- group_by(subset(pf, !is.na(gender)), age, gender) 
pf.fc_by_age_gender <- summarise(age_gender_groups, 
                                 mean_friend_count=mean(friend_count),
                                 median_friend_count=median(friend_count),
                                 n=n()) 
pf.fc_by_age_gender <- arrange(pf.fc_by_age_gender, age) 
head(pf.fc_by_age_gender) 
tail(pf.fc_by_age_gender) 
#Solving with chained commands
pf.fc_by_age_gender1 <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count=mean(friend_count),
            median_friend_count=median(friend_count),
            n=n()) %>%
  ungroup() %>% #Summarise removes one layer of grouping when it runs, gender is removed, so ungroup 1 more time to remove age layer
  arrange(age)
head(pf.fc_by_age_gender1)
tail(pf.fc_by_age_gender1)

# Create a line graph showing the
# median friend count over the ages
# for each gender. Be sure to use
# the data frame you just created,
# pf.fc_by_age_gender.
ggplot(aes(x = age, y = median_friend_count), 
              data = subset(pf.fc_by_age_gender1)) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:
Long format, there is a row for each gender 
Wide format, make 2 columns for gender (so half the rows than Long)
***

### Reshaping Data
Notes: Functions used
  speread
  dcast/acast
  melt
```{r}
#install.packages('reshape2')
library(reshape2)
library(tidyr)
#using dplyr-tidyr
spread(subset(pf.fc_by_age_gender, select = c('gender', 'age', 'median_friend_count')), gender, median_friend_count)

#using reshape
pf.fc_by_age_gender.wide <- dcast( #dcast for dataframe, acast for vector/array
                          pf.fc_by_age_gender, 
                          age ~ gender,
                          value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)
#melt function converts from wide to long format
```


***

### Ratio Plot
Notes:
# Plot the ratio of the female to male median
# friend counts using the data frame
# pf.fc_by_age_gender.wide.

# Think about what geom you should use.
# Add a horizontal line to the plot with
# a y intercept of 1, which will be the
# base line. Look up the documentation
# for geom_hline to do that. Use the parameter
# linetype in geom_hline to make the
# line dashed.

```{r Ratio Plot}
ggplot(aes(x=age, y=female/male), data=pf.fc_by_age_gender.wide) +
  geom_line() +
  geom_hline(yintercept = 1, alpha=0.3, linetype=2)

```

***

### Third Quantitative Variable
Notes:
year_joined based on tenure (in days)
```{r Third Quantitative Variable}
pf$year_joined <- floor(2014 - pf$tenure/365)
summary(pf$year_joined)
table(pf$year_joined)

```

***

### Cut a Variable
Notes:
# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

# Note that a parenthesis means exclude the year and a
# bracket means include the year.
```{r Cut a Variable}
pf$year_joined.bucket=cut(pf$year_joined, 
                          c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket)

#cut samples
Z <- stats::rnorm(10000)
table(cut(Z, breaks = -6:6))
sum(table(cut(Z, breaks = -6:6, labels = FALSE)))
(graphics::hist(Z, breaks = -6:6, plot = FALSE)$counts)
table(cut(Z, breaks=3*(-2:5), right=FALSE))
```

***

### Plotting it All Together
Notes:
# Create a line graph of friend_count vs. age
# so that each year_joined.bucket is a line
# tracking the median user friend_count across
# age. This means you should have four different
# lines on your plot.

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA='ifany')
ggplot(aes(x=age, y=friend_count), 
           data=subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=median)

```

***

### Plot the Grand Mean
Notes:
# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.

# (2) Exclude any users whose year_joined.bucket is NA.

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash
```{r Plot the Grand Mean}
ggplot(aes(x=age, y=friend_count), 
           data=subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=mean, linetype=2)
#The grand mean line shows that much of the pattern is due to recent joinees
```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(subset(pf, tenure>=1), summary(friend_count/tenure))
```

***

### Friendships Initiated
Notes:

What is the median friend rate?
0.22
What is the maximum friend rate?
417
```{r Friendships Initiated}

```

***

### Bias-Variance Tradeoff Revisited
Notes:
# Create a line graph of mean of friendships_initiated per day (of tenure)
# vs. tenure colored by year_joined.bucket.

# You need to make use of the variables tenure,
# friendships_initiated, and year_joined.bucket.

# You also need to subset the data to only consider user with at least
# one day of tenure.

```{r Bias-Variance Tradeoff Revisited}
#People with more tenure initiate less friendships
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)
#Smoothing the curve to remove noise (by binning the x axis differently)
ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:
Discreteness in the distribution, some prices have a distribution while others don't
```{r Histograms Revisited}
yo <- read.csv('yogurt.csv')
str(yo)
summary(yo)
yo$id <- factor(yo$id)
str(yo)
ggplot(data = yo, aes(x = price)) + 
  geom_histogram(binwidth = 1) 
qplot(data=yo, x=price, fill='#F79420')
```

***

### Number of Purchases
Notes:
Discreteness can be observed by looking at unique prices, 20 are there

# Create a new variable called all.purchases,
# which gives the total counts of yogurt for
# each observation or household.

# One way to do this is using the transform
# function. You can look up the function transform
# and run the examples of code at the bottom of the
# documentation to figure out what it does.

# The transform function produces a data frame
# so if you use it then save the result to 'yo'!

```{r Number of Purchases}
length(unique(yo$price))
table(yo$price) #also gives an idea of distribution
yo <- transform(yo, all.purchases=(strawberry + blueberry + pina.colada + mixed.berry + plain))
summary(yo$all.purchases)
```

***

### Prices over Time
Notes:
# Create a scatterplot of price vs time.

# This will be an example of a time series plot.

# Resolve overplotting issues by using
# techniques you learned in Lesson 4.

Most common prices seem to be increasing over time. some lower prices may be due to sales/coupons
```{r Prices over Time}
qplot(data=yo, x=all.purchases, fill='#F79420', binwidth=1)
ggplot(aes(x=time, y=price), data=yo) +
  geom_jitter(alpha=0.25, shape=21, fill = I('#F79420'))
```

***

### Sampling Observations
Notes:
This data has Multiple observations of the same unit, work with a sample of those units to display raw data easily for that sample. Look at small set of households in more detail. 
***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(2324)
sample.ids <- sample(levels(yo$id), 16)
#%in% selects only those ids, in the sample.ids
ggplot(aes(x=time, y=price), data=subset(yo, id %in% sample.ids)) + 
  facet_wrap( ~ id) +
  geom_line() + 
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:
Categorical variables create faceted histograms,
Continuous variables create scatter plots
```{r}
library(GGally)
theme_set(theme_minimal(20))
set.seed(1836)
pf_subset <- pf[,c(2:15)] #only variable indexes 2-15, 14 variables
names(pf)
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset),1000),])
  #use axisLabels = 'internal' to have labels inside the plot
```
***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

