Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
qplot(x=age, y=friend_count, data=pf)
qplot(age, friend_count, data=pf)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:
20 points makeup 1 black dot, reducing the overplotting
```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_point(alpha=1/20) + #20 points makeup 1 black dot, reducing the overplotting
    xlim(13, 90) #adds a new layer to the plot
summary(pf$age)
```

***

### Overplotting
Notes:
jitter makes the plot truely representing the age with more dispersed distribution

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_jitter(alpha=1/20) + #20 points makeup 1 black dot, reducing the overplotting
    xlim(13, 90) #adds a new layer to the plot
summary(pf$age)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:
if geom_jitter is needed then some care is needed: for people with 0 friend count, if noise is added then there might be negative values for friend count and square root of imaginary numbers. Hence pass jitter for position only and set height = 0
```{r Coord_trans()}

ggplot(aes(x=age, y=friend_count), data=pf) +
    geom_point(alpha=1/20, position=position_jitter(h=0)) + #20 points makeup 1 black dot, reducing the overplotting
    xlim(13, 90) +
    coord_trans( y='sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}

ggplot(aes(x=age, y=friendships_initiated), data=pf) +
    #geom_point()
    #geom_jitter(alpha=1/10)
    geom_point(alpha=1/10, position='jitter')

ggplot(aes(x=age, y=friendships_initiated), data=pf) +
    #geom_point()
    #geom_jitter(alpha=1/10)
    geom_jitter(alpha=1/10, position=position_jitter(h=0))+
    coord_trans(y='sqrt')


```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:
DPLYR - splits up a DF and apply a function to some parts of the data
common functions filter | groupby | mutate | arrange

```{r Conditional Means}
library(dplyr)
age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n= n()) # number of users in the group
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)

#pf.fc_by_age <- pf %>%
#  group_by(age) %>% #This allows to chain functions on the dataset
#  summarise(age_groups,
#          friend_count_mean = mean(friend_count),
#          friend_count_median = median(friend_count),
#          n= n()) %>%
#  arrage(age)
#head(pf.fc_by_age)
```

Create your plot!
# Plot mean friend count vs. age using a line graph.
# Be sure you use the correct variable names
# and the correct data frame. You should be working
# with the new data frame created from the dplyr
# functions. The data frame is called 'pf.fc_by_age'.

# Use geom_line() rather than geom_point to create
# the plot. You can look up the documentation for
# geom_line() to see what it does.
```{r Conditional Means Plot}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  geom_line()
```
Spike is seen at age 69, for older ages, the estimates are highly variable
***

### Overlaying Summaries with Raw Data
Notes:
Example for calculating percentiles
http://www.r-tutor.com/elementary-statistics/numerical-measures/percentile
To zoom in, the code should use the coord_cartesian(xlim = c(13, 90)) layer rather than xlim(13, 90) layer.

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x=age, y=friend_count), data=pf) +
    geom_point(alpha=1/20, position=position_jitter(h=0),
               color='orange') + 
    xlim(13, 90) +
    coord_trans( y='sqrt') +
  geom_line(stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.1),
            linetype=2, color='blue') +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.9),
            linetype=2, color='green')
#With coord_cartesian
ggplot(aes(x=age, y=friend_count), data=pf) +
  coord_cartesian(xlim=c(13, 70), ylim=c(0,1000)) +
    geom_point(alpha=1/20, position=position_jitter(h=0),
               color='orange') + 
  geom_line(stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.1),
            linetype=2, color='blue') +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.9),
            linetype=2, color='green')
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:
>+-0.3 < +-0.5 means moderate
>+-0.5 is a strong correlation
```{r Correlation}
head(pf$age)
cor.test(pf$age, pf$friend_count, method='pearson')
with(pf, cor.test(age, friend_count, method='pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:
Pearson is the default method 
```{r Correlation on Subsets}
with(subset(pf, age<70) , cor.test(age, friend_count))

with(subset(pf, age<70) , cor.test(age, friend_count, method='spearman'))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:
# Create a scatterplot of likes_received (y)
# vs. www_likes_received (x). Use any of the
# techniques that you've learned so far to
# modify the plot.
```{r}
ggplot(aes(x=pf$www_likes_received, y=pf$likes_received), data=pf) +
    geom_point()
``` 

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x=pf$www_likes_received, y=pf$likes_received), data=pf) +
    geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method='lm', color='red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
cor.test(pf$likes_received, pf$www_likes_received)
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:
Argument matching (when not providing them by name) in R is a bit complex.

First, arguments (or parameters) can be matched by name. If a parameter matches exactly, it is "removed" from the argument list and the remaining unnamed arguments are matched in the order that they are listed in the function definition.

R does the following to match arguments... 

checks for exact match of named argument
checks for a partial match of the argument
checks for a positional match
If R does not find a match for a parameter, it typically throws an "unused" parameter error.

Type str(functionName) to find the order of the parameters and learn more about the parameters of an R function. 

```{r More Caution With Correlation}
library(alr3)
data(Mitchell)
#?Mitchell
```

Create your plot!
# Create a scatterplot of temperature (Temp)
# vs. months (Month).
```{r Temp vs Month}
head(Mitchell)
ggplot(aes(x=Temp, y=Month), data=Mitchell) + 
  geom_point()

qplot(data=Mitchell, Month, Temp)

```
***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
cor.test(Mitchell$Month, Mitchell$Temp)
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(x=Month, y=Temp), data=Mitchell) + 
  geom_point() + 
  #scale_y_continuous("Months") + 
  scale_x_continuous(breaks=seq(0, 203, 12))
```

***

### A New Perspective

What do you notice?
Response:
Zoom the graph and stretch so that width is 2wice the height, see sinusoidal

The cor and cor.test functions determine the strength of a linear relationship, but they may miss other relationships in the data.

Watch the solution video and check out the Instructor Notes!
Notes:
The dcor.ttest() function in the energy package implements a non-parametric test of the independence of two variables. While the Mitchell soil dataset is too coarse to identify a significant dependency between "Month" and "Temp", we can see the difference between dcor.ttest and cor.test through other examples, like the following:
```{r }
ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell) + 
  geom_point() 

x <- seq(0, 4*pi, pi/20)
y <- cos(x)
qplot(x = x, y = y)
#dcor.ttest(x, y)
```
***

### Understanding Noise: Age to Age Months
Notes:
# Create a new variable, 'age_with_months', in the 'pf' data frame.
# Be sure to save the variable in the data frame rather than creating
# a separate, stand-alone variable. You will need to use the variables
# 'age' and 'dob_month' to create the variable 'age_with_months'.

```{r Understanding Noise: Age to Age Months}
pf$age_with_months <- pf$age + (1-pf$dob_month/12)
```

***### Age with Months Means
# Create a new data frame called
# pf.fc_by_age_months that contains
# the mean friend count, the median friend
# count, and the number of users in each
# group of age_with_months. The rows of the
# data framed should be arranged in increasing
# order by the age_with_months variable.
```{r Age with Months Means}
library(dplyr)
age_groups_mnth <- group_by(pf, age_with_months)
pf.fc_by_age_months <- summarise(age_groups_mnth,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n= n()) # number of users in the group
pf.fc_by_age_months <- arrange(pf.fc_by_age_months, age_with_months)
head(pf.fc_by_age_months)

```

Programming Assignment
# Create a new scatterplot showing friend_count_mean
# versus the new variable, age_with_months. Be sure to use
# the correct data frame (the one you create in the last
# exercise) AND subset the data to investigate
# users with ages less than 71.
Create a NOISY Plot
```{r Programming Assignment}
ggplot(aes(x=age_with_months, y=friend_count_mean), 
       data=subset(pf.fc_by_age_months, age_with_months<71)) +
  geom_line()
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:
Finer bin choices (with age in months) = more noise 
Bias Variance Tradeoff
```{r Smoothing Conditional Means}
p1 <- ggplot (aes(x=age, y=friend_count_mean),
              data=subset(pf.fc_by_age, age<71)) + 
  geom_line() + 
  geom_smooth() #The smoother in blue

p2 <- ggplot(aes(x=age_with_months, y=friend_count_mean), 
       data=subset(pf.fc_by_age_months, age_with_months<71)) +
  geom_line() + 
  geom_smooth() #The smoother in blue, you miss some sudden peaks
#rounding makes the curve smoother
p3 <- ggplot(aes(x=round(age/5)*5, y=friend_count),
             data=subset(pf, age<71)) +
  geom_line(stat='summary', fun.y=mean)

library(gridExtra)
grid.arrange(p2, p1, p3, ncol=1)
```

***

### Which Plot to Choose?
Notes:
Don't choose in exploratory analysis
***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

